---
output:
  github_document
    #fig_width: 7
    #fig_height: 5 # default is 5
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# paintr

<!-- badges: start -->
[![R-CMD-check](https://github.com/coatless-rpkg/paintr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/coatless-rpkg/paintr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of `paintr` is to draw different _R_ data structures on graphs.

> [!NOTE]
>
> A previous version of the package was called `drawr`; however, another package
> with the same name was published on CRAN. As a result, the package was renamed
> to `paintr`.

## Installation

You can install the development version of drawr from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("coatless-rpkg/paintr")
```

## Design

The package is designed to take advantage of base R graphics alongside `ggplot2`.
We're providing two different implementations for each system under the
naming scheme of:

- `paint_*()`: base R graphics
- `gpaint_*()`: `ggplot2`

## Example

Take for instance we have a matrix that looks like so: 

```{r}
mat_3x5 = matrix(
  c(
   1, NA,    3,   4,  NaN, 
  NA,  7,    8,  -9,  10, 
 -11, 12, -Inf, -14,  NA
 ),
 ncol = 5, byrow = TRUE)

mat_3x5
```

What if we wanted to see the contents laid out with their indices or specific cells
highlighted?

```{r}
#| label: base-example
#| results: 'markup'
# Load the library
library(paintr)

# Graphic of matrix data structure using base R graphics
paint_matrix(mat_3x5)
# Show the cell indices
paint_matrix(mat_3x5, show_indices = "cell")
# Show all indices
paint_matrix(mat_3x5, show_indices = "all")
# Highlight cells over a specific value
paint_matrix(mat_3x5, highlight_area = mat_3x5 > 4)
```

We can achieve similar results with the `ggplot2` function. 

```{r}
#| label: ggplot2-example
#| results: 'markup'
# Graphic of matrix data structure using base R graphics
gpaint_matrix(mat_3x5)
# Highlight cells in specific columns
gpaint_matrix(mat_3x5, 
             show_indices = c("row", "column"),
             highlight_area = highlight_columns(mat_3x5, columns = 2:4))
```



