---
title: "Getting started with paintr"
format:
  html:
    minimal: true
    embed-resources: true
vignette: >
  %\VignetteIndexEntry{Getting started with paintr}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5,
                      dpi = 96, out.width = "100%")
```

```{r}
library(paintr)
```

## What paintr draws

When you teach R, the hard part is rarely the arithmetic. It is helping someone
*see* the shape of the thing they are working with: that a vector is a row of
boxes, that a data frame is a stack of columns that happen to line up, that an
array is a pile of matrices. paintr draws exactly those pictures.

```{r}
#| fig-alt: "A numeric vector of six values, 4, 8, 15, 16, 23, and 42, drawn as a column of cells;
#|   under each cell is its positional accessor, [1] through [6], the expression you type to reach
#|   that value."
paint_vector(c(4, 8, 15, 16, 23, 42))
```

Every picture is the same idea. Each value sits in its own cell, and under each
cell paintr writes **the expression you would type to reach it**. The label
under the third box above is `[3]`, because `x[3]` is how you would pull that
value out. That is the whole teaching trick: the drawing and the code are the
same object.

## Two backends: `paint_` and `gpaint_`

Every structure has two painters. The `paint_*()` functions draw in base
graphics, straight to the device. The `gpaint_*()` functions return a **ggplot
object**, so you can print it, save it with `ggsave()`, or drop it into a
patchwork.

```{r}
#| fig-alt: "The same six values, 4, 8, 15, 16, 23, and 42, drawn as a column of cells with the
#|   accessors [1] through [6] beneath them, this time rendered as a ggplot object rather than
#|   base graphics; the picture is identical."
gpaint_vector(c(4, 8, 15, 16, 23, 42))
```

The two draw the same picture. One honest caveat about the ggplot version: its
panel is a single custom grob, not a layer of geoms. That is what lets paintr
defer text fitting to draw time and paint the grey trailing digits you will see
below, but it also means `ggplot_build()` sees an empty layer, and adding
`+ scale_fill_*()` has no effect. Reach for `gpaint_*()` when you want a ggplot
in your hand, and `paint_*()` otherwise. The rest of this tour uses `paint_*()`.

## Vectors

A vector is a one-dimensional run of values, and that is how it draws.

```{r}
#| fig-alt: "A numeric vector of six values drawn as a column of cells holding -3, 5, NA, Inf, 2,
#|   and 1; the missing value shows as NA and the infinite value as Inf, and each cell carries its
#|   positional accessor [1] through [6]."
paint_vector(c(-3, 5, NA, Inf, 2, 1))
```

Two options earn their keep here. `layout = "horizontal"` turns the column on
its side, and `show_indices` writes the position labels `[1] [2] ...` either
`"inside"` each cell or `"outside"` the row.

```{r}
#| fig-alt: "The numbers 1 through 6 drawn as a single horizontal row of cells instead of a column,
#|   with the positional labels [1] through [6] printed outside the row beneath each cell."
paint_vector(1:6, layout = "horizontal", show_indices = "outside")
```

Give the vector names and it draws each name as the accessor you would type,
`["mon"]` rather than a bare `mon`, because `x["mon"]` is how you reach the
value. The through-line holds: the label under every cell is still exactly the
expression you type. Set `show_names = FALSE` to fall back to `[1] [2]`.

```{r}
#| fig-alt: "A named numeric vector of four values, 12, 19, 3, and 8, drawn as a column of cells;
#|   each cell is labelled with its name as a character accessor, [\"mon\"], [\"tue\"], [\"wed\"],
#|   and [\"thu\"], the expression you type to reach that value."
paint_vector(c(mon = 12, tue = 19, wed = 3, thu = 8))
```

## Matrices

A matrix is a grid, so it draws as one. `paint_matrix()` also handles a 2-D
array or a 2-D table.

```{r}
#| fig-alt: "The numbers 1 through 15 drawn as a 3 by 5 grid of cells filled column by column, so
#|   the first column reads 1, 2, 3 down and the last reads 13, 14, 15."
paint_matrix(matrix(1:15, nrow = 3))
```

The one option you will reach for most is `show_indices = "cell"`, which labels
every cell with its `[row, column]` subscript. This is the picture that makes
`m[2, 4]` finally click.

```{r}
#| fig-alt: "The same 3 by 5 grid of the numbers 1 through 15, now with each cell labelled by its
#|   [row, column] subscript, from [1, 1] in the top-left corner to [3, 5] in the bottom-right."
paint_matrix(matrix(1:15, nrow = 3), show_indices = "cell")
```

If the matrix has dimnames, they draw in the margins; `show_dimnames` takes a
character vector over `"none"`, `"row"`, `"column"`, and `"all"` to choose which.

## Data frames

A data frame is the workhorse, and it gets the richest default picture: a header
row of column names, the type of each column beneath it, and a row-name gutter
whenever the row names are meaningful.

```{r}
#| fig-height: 5
#| fig-alt: "The first six rows of the mtcars data frame drawn as a table: a header row of bare
#|   column names such as mpg, cyl, and disp, the column type beneath each, and a row-name gutter
#|   down the left labelling each row as an accessor, [\"Mazda RX4\", ] through [\"Valiant\", ];
#|   numbers show three significant figures with any trailing digits in grey."
paint_data_frame(head(mtcars))
```

Notice the numbers. paintr shows three significant figures by default, and draws
any trailing insignificant digits in **grey**, so a reader can see the true
value without being distracted by noise. `mtcars` has real row names, so the
gutter appears; a fresh `iris` would not draw one. Toggle the extra rows with
`show_names`, `show_types`, and `show_rownames`. `paint_df()` is a shorter alias.

## Lists

A list is the flexible container: its elements can have different lengths and
different types. paintr draws each element as a column.

```{r}
#| fig-alt: "A list of three elements drawn as three side-by-side columns headed $id, $tags, and
#|   $ok; id holds 1, 2, 3, tags holds a and b, and ok holds TRUE, FALSE, NA. The tags column is
#|   one cell shorter, leaving a ragged bottom edge because the elements do not share a length."
paint_list(list(id = 1:3, tags = c("a", "b"), ok = c(TRUE, FALSE, NA)))
```

Here is the idea worth pausing on. A data frame is just a list whose elements
all share a length, so when the elements *do* line up, the list draws as a clean
rectangle. Break one length and the rectangle breaks with it, ragged edge and
all. That ragged edge is the difference between a list and a data frame, drawn
literally.

paintr does not draw nesting: a list inside a list collapses to a grey
`<list [2]>` token rather than recursing. Use `show_indices = "cell"` to label
each cell with its `[[j]][i]` accessor.

## Arrays

An array of three or more dimensions is a stack of matrices, and `paint_array()`
lays that stack out as a row of blocks sharing one font size. (A 2-D array is
just a matrix, and draws like one.)

```{r}
#| fig-width: 7.7
#| fig-height: 6.6
#| fig-alt: "The numbers 1 through 24 drawn as a 2 by 3 by 4 array: four side-by-side blocks, each
#|   a 2 by 3 matrix slice sharing one font size and titled , , 1 through , , 4. Every cell is
#|   labelled with its full three-number subscript, from [1, 1, 1] to [2, 3, 4]."
paint_array(array(1:24, dim = c(2, 3, 4)), show_indices = "cell")
```

Each block is one slice; the label under a cell is its full subscript, `[2, 3, 4]`,
a genuine 3-D index into the array. `show_dimnames` works as it does for a
matrix, with an added `"slice"` lane. A genuinely huge array elides its middle,
the same way the other structures do once they pass their size caps.

When a stack has more slices than sit comfortably in one row, `slices_per_row`
wraps the blocks onto a grid -- the same array, laid out to stay readable. Each
block keeps its own honest slice title, so wrapping is just reading order.

```{r}
#| fig-height: 4.5
#| fig-alt: "The same 2 by 3 by 4 array, now with slices_per_row = 2, so the four slice blocks are
#|   laid out in a two-by-two grid instead of a single row. The blocks are titled , , 1 through
#|   , , 4 in left-to-right, top-to-bottom reading order, and are larger and easier to read than in
#|   one cramped row."
paint_array(array(1:24, dim = c(2, 3, 4)), slices_per_row = 2)
```

## Picking a legible device size

Most defaults fit a normal plot window, but a large structure drawn in full can
shrink its text below the point where it is readable. `paint_size()` tells you
the device size that keeps everything legible, in a form you can paste straight
into a `png()` call or a knitr chunk header.

```{r}
paint_size(head(mtcars))
```

The answer is in inches by default; pass `units = "px"` for a `png()` at a given
`dpi`. When a picture would be too small, paintr warns and quotes you these exact
numbers.

## Where next

That is the whole tour: five structures, two backends, one consistent picture.
From here:

- `vignette("highlighting", package = "paintr")` shows how to spotlight rows,
  columns, and individual cells to draw the eye where a lesson needs it.
- `vignette("base-vs-ggplot2", package = "paintr")` digs into the `paint_` and
  `gpaint_` split and when each backend pays off.
- `vignette("teaching", package = "paintr")` puts these pictures to work in
  actual lesson material.
